# Accesing the data

## Before downloading datasets

If you are going to download data, you have to read the [Code of Conduct](https://docs.tradestatistics.io/index.html#code-of-conduct) first.

## API

The advantage of the API over https download is that you can filter what to obtain and also access some additional tables.

To obtain exactly the same data as with compressed files, please refer to \@ref(yrpc-year-reporter-partner-and-product-code).

If you use R you'll need `jsonlite` and `dplyr` packages.

```{r message=FALSE, warning=FALSE}
library(jsonlite)
```

These packages are also useful:

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
```

### Available tables

```{r, warning=FALSE, message=FALSE, cache = TRUE}
as_tibble(fromJSON("https://api.tradestatistics.io/tables"))
```

### Metadata

```{r, warning=FALSE, message=FALSE, cache = TRUE}
## Countries (no filter)
rda_countries <- "countries.rda"

if (!file.exists(rda_countries)) {
  countries <- as_tibble(fromJSON(
    "https://api.tradestatistics.io/countries"
  ))

  save(countries, file = rda_countries, compress = "xz")

  countries
} else {
  load(rda_countries)

  countries
}

## Products (no filter)
rda_products <- "products.rda"

if (!file.exists(rda_products)) {
  products <- as_tibble(fromJSON(
    "https://api.tradestatistics.io/products"
  ))

  save(products, file = rda_products, compress = "xz")

  products
} else {
  load(rda_products)

  products
}
```

Please notice that these tables include some aliases. 

`countries` includes some meta-codes, `c-xx` where `xx` must the first two letters of a continent and `all`, this is:

```{r, echo = FALSE}
library(knitr)

countries %>%
  filter(str_sub(country_iso, 1, 2) == "c-" | country_iso == "all") %>%
  select(country_iso, country_name_english) %>%
  rename(
    `Alias` = country_iso,
    `Meaning` = country_name_english
  ) %>%
  kable()
```

`products` also includes some meta-codes, `xx` for the first two digits of a code and those digits are the product group and `all`, this is:

```{r, echo = FALSE}
products %>%
  filter(str_length(product_code) %in% 2:3) %>%
  select(product_code, product_fullname_english) %>%
  rename(
    `Alias` = product_code,
    `Meaning` = product_fullname_english
  ) %>%
  kable()
```

### API parameters

The tables provided withing our API contain at least one of these fields:

* Year (`y`) 
* Reporter ISO (`r`)
* Partner ISO (`p`)
* Product Code (`c`)

The most detailed table is `yrpc` that contains all bilateral flows at product level.

With respect to `y` you can pass any integer contained in $[1962,2018]$.

Both `r` and `p` accept any valid ISO code or alias contained in the [countries](https://api.tradestatistics.io/countries) table. For example, both `chl` (valid ISO code) and `c-am` (continent Americas, an alias) are valid API filtering parameters.

`c` takes any valid product code or alias from the [products](https://api.tradestatistics.io/products). For example, both `0101` (valid HS product code) and `01` (valid HS group code) are valid API filtering parameters.

By default the API takes `c = "all"` by default.

You can always skip `c`, but `y`, `r` and `p` are requiered to return data.

### Available reporters

The only applicable filter is by year.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Available reporters (filter by year)
as_tibble(fromJSON(
  "https://api.tradestatistics.io/reporters?y=2018"
))
```

### YRPC (Year, Reporter, Partner and Product Code)

The applicable filters here are year, reporter, partner, product code and (optionally) product code length.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Reporter - Partner - Product Code (filter by year, reporter and partner)

## file to store the query results
rda_1962 <- "api_data_1962.rda"

## filter by product code length (parameter `l`)
if (!file.exists(rda_1962)) {
  yrpc_1 <- as_tibble(fromJSON(
    "https://api.tradestatistics.io/yrpc?y=1962&r=usa&p=all"
  ))

  save(yrpc_1, file = rda_1962, compress = "xz")

  yrpc_1
} else {
  load(rda_1962)

  yrpc_1
}

## filter by product group (parameter `c`)
yrpc_2 <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrpc?y=2018&r=chl&p=arg&c=01"
))

yrpc_2
```

Some columns requiere an explanation:

* `product_code`: HS07 product codes (e.g. according to the \code{products} table within this package, 0101 stands for "Horses, etc.")
* `product_code_length`: How many digits does `product_code` contain, this can be useful to filter by depth when using HS codes (HS 6 digits is a more detailed version of HS 4 digits, and therefore you don't have to sum both or you'll be counting exports/imports twice)
* `group_code`: International categorization of group products defined after product ID
* `group_name`: English name corresponding to `group_id`
* `export_value_usd`: Exports measured in nominal United States Dollars (USD)
* `import_value_usd`: Imports measured in nominal United States Dollars (USD)
* `export_value_usd_percentage_change_1_year`: Nominal increase/decrease in exports measured as percentage with respect to last year
* `export_value_usd_percentage_change_5_years`: Nominal increase/decrease in exports measured as percentage with respect to five years ago
* `export_value_usd_change_1_year`: Nominal increase/decrease in exports measured in USD with respect to last year
* `export_value_usd_change_5_years`: Nominal increase/decrease in exports measured in USD with respect to five years ago

### YRC (Year, Reporter and Product Code)

The only applicable filter is by year, reporter, product code and (optionally) product code length.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Reporter - Product Code (filter by year and reporter)

## filter by reporter ISO (parameter `r`)
yrc_1 <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrc?y=2018&r=chl"
))

yrc_1

## filter by reporter alias (also parameter `r`)
yrc_2 <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrc?y=2018&r=c-am"
))

yrc_2
```

Here the `export_rca*` and `import_rca*` fields contain the Revealed Comparative Advantage (RCA) of an exported product with respect to all the products with the same number of digits. The definition of RCA is detailed on [Open Trade Statistics Documentation](https://tradestatistics.github.io/documentation/).

### YRP (Year, Reporter and Partner)

The only applicable filter is by year, reporter and partner.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Reporter - Partner (filter by year, reporter and partner)
yrp <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrp?y=2018&r=chl&p=arg"
))
```

### YC (Year and Product Code)

The only applicable filter is by year, product and (optionally) product code length.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Product Code (filter by year)
yc <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yc?y=2018&c=0101"
))
```

Let's explore the first rows of `yr`:

```{r}
yc
```

Here some fields deserve an explanation:

* `pci_4_digits_product_code`: Product Complexity Index (PCI) which is detailed on [Open Trade Statistics Documentation](https://tradestatistics.github.io/documentation/). This index is built by using just four digits product codes.
* `pci_6_digits_product_code`: Similar to the previous field but built by using just six digits product codes.
* `pci_rank_4_digits_product_code`: The rank of a product given its PCI (e.g. the highest PCI obtains the #1)
* `pci_rank_4_digits_product_code_delta_1_year`: How many places a country increased or decreased with respect to last year

#### YR (Year and Reporter)

The only applicable filter is by year and reporter.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
## Year - Reporter (filter by year and reporter)
yr <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yr?y=2018&r=chl"
))
```

Let's explore the first rows of `yr`:

```{r}
yr
```

Some fields here require more detail:

* `eci_4_digits_product_code`: Economic Complexity Index (ECI) which is detailed on [Open Trade Statistics Documentation](https://tradestatistics.github.io/documentation/). This index is built by using just four digits product codes.
* `eci_rank_4_digits_product_code`: The rank of a country given its ECI (e.g. the highest ECI obtains the #1)
* `eci_rank_4_digits_product_code_delta_1_year`: How many places a country increased or decreased with respect to last year

### Country rankings

The only applicable filter is by year.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Country rankings (filter by year)
country_rankings <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/country_rankings?y=2018"
))
```

### Product rankings

The only applicable filter is by year.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Product rankings (filter by year)
product_rankings <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/product_rankings?y=2018"
))
```

## R Package

To ease API using, we provide an [R Package](https://ropensci.github.io/tradestatistics/). This package is a part of [ROpenSci](https://ropensci.org/) and its documentation is available on a separate [pkgdown site](https://ropensci.github.io/tradestatistics/).

## Dashboard (beta)

To ease API using, we provide a [Shiny Dashboard](https://shiny.tradestatistics.io/) that is still under improvements.

## RDS datasets

Please check the [md5sums](https://docs.tradestatistics.io/direct-download/md5sums.txt) to verify data integrity after downloading.

```{r, echo = FALSE, message = F}
library(readr)
library(tidyr)

files <- list.files("./docs/direct-download", pattern = "rds", full.names = TRUE)
urls <- str_replace(files, "./docs/direct-download/", "https://docs.tradestatistics.io/direct-download/")

md5sums <- read_fwf("./docs/direct-download/md5sums.txt", fwf_widths(c(32, 15), c("md5sum", "file"))) %>% mutate(file = str_trim(file))

tibble(
  `Year` = paste0("[", 1962:2018, "](", urls, ")"),
  `Last updated` = "2020-02-11",
  `File size (MB)` = round(file.size(files) / 1024^2, 2),
  `MD5 sum` = md5sums$md5sum
) %>%
  kable()
```
