# Accesing the data

## Before downloading datasets

If you are going to download data, you have to read the [Code of Conduct](https://docs.tradestatistics.io/index.html#code-of-conduct) first.

## API

The advantage of the API over https download is that you can filter what to obtain and also access some additional tables.

To obtain exactly the same data as with compressed files, please refer to \@ref(yrpc-year-reporter-partner-and-product-code).

If you use R you'll need `jsonlite` and `dplyr` packages.

```{r message=FALSE, warning=FALSE}
library(jsonlite)
```

These packages are also useful:

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
```

### Available tables

```{r, warning=FALSE, message=FALSE, cache = TRUE}
as_tibble(fromJSON("https://api.tradestatistics.io/tables"))
```

### Metadata

```{r, warning=FALSE, message=FALSE, cache = TRUE}
## Countries (no filter)
rda_countries <- "countries.rda"

if (!file.exists(rda_countries)) {
  countries <- as_tibble(fromJSON(
    "https://api.tradestatistics.io/countries"
  ))

  save(countries, file = rda_countries, compress = "xz")

  countries
} else {
  load(rda_countries)

  countries
}

## Products (no filter)
rda_products <- "products.rda"

if (!file.exists(rda_products)) {
  products <- as_tibble(fromJSON(
    "https://api.tradestatistics.io/products"
  ))

  save(products, file = rda_products, compress = "xz")

  products
} else {
  load(rda_products)

  products
}
```

Please notice that these tables include some aliases. 

`countries` includes some meta-codes, `c-xx` where `xx` must the first two letters of a continent and `all`, this is:

```{r, echo = FALSE}
library(knitr)

countries %>%
  filter(str_sub(country_iso, 1, 2) == "c-" | country_iso == "all") %>%
  select(country_iso, country_name_english) %>%
  rename(
    `Alias` = country_iso,
    `Meaning` = country_name_english
  ) %>%
  kable()
```

`products` also includes some meta-codes, `xx` for the first two digits of a code and those digits are the product group and `all`, this is:

```{r, echo = FALSE}
products %>%
  filter(str_length(product_code) %in% 2:3) %>%
  select(product_code, product_fullname_english) %>%
  rename(
    `Alias` = product_code,
    `Meaning` = product_fullname_english
  ) %>%
  kable()
```

### API parameters

The tables provided withing our API contain at least one of these fields:

* Year (`y`) 
* Reporter ISO (`r`)
* Partner ISO (`p`)
* Product Code (`c`)

The most detailed table is `yrpc` that contains all bilateral flows at product level.

With respect to `y` you can pass any integer contained in $[1962,2018]$.

Both `r` and `p` accept any valid ISO code or alias contained in the [countries](https://api.tradestatistics.io/countries) table. For example, both `chl` (valid ISO code) and `c-am` (continent Americas, an alias) are valid API filtering parameters.

`c` takes any valid product code or alias from the [products](https://api.tradestatistics.io/products). For example, both `0101` (valid HS product code) and `01` (valid HS group code) are valid API filtering parameters.

By default the API takes `c = "all"` by default.

You can always skip `c`, but `y`, `r` and `p` are requiered to return data.

### Available reporters

The only applicable filter is by year.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Available reporters (filter by year)
as_tibble(fromJSON(
  "https://api.tradestatistics.io/reporters?y=2018"
))
```

### YRPC (Year, Reporter, Partner and Product Code)

The applicable filters here are year, reporter, partner and product code.

```{r yrpc, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Reporter - Partner - Product Code

yrpc_1 <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrpc?y=1962&r=usa&p=mex&c=8703"
))

yrpc_1
```

Columns definition:

* `reporter_iso`: Official ISO-3 code for the reporter (e.g. the country that reports X dollars in exports/imports from/to country Y)
* `partner_iso`: Official ISO-3 code for the partner
* `product_code`: Official Harmonized System rev. 2007 (HS07) product code (e.g. according to the \code{products} table in the API, 8703 stands for "Motor cars and other motor vehicles; principally designed for the transport of persons (other than those of heading no. 8702), including station wagons and racing cars")
* `export_value_usd`: Exports measured in nominal United States Dollars (USD)
* `import_value_usd`: Imports measured in nominal United States Dollars (USD)

### YRPC-GA (Year - Reporter - Partner - Product Code, Group Code Aggregated)

The applicable filters here are year, reporter, partner and group code. Here the group code is just an aggregation over product code.

```{r yrpc-ga, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Reporter - Partner - Product Code, Group Code Aggregated

yrpc_ga <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrpc-ga?y=1962&r=usa&p=mex&g=87"
))

yrpc_ga
```

Columns definition:

* `group_code`: Official Harmonized System rev. 2007 (HS07) group code (e.g. according to the \code{products} table in the API, 87 stands for "Vehicles; other than railway or tramway rolling stock, and parts and accessories thereof")

### YRPC-CA (Year - Reporter - Partner - Product Code, Community Code Aggregated)

The applicable filters here are year, reporter, partner and community code. Here the community code is just an aggregation over both product code and group code.

```{r yrpc-ca, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Reporter - Partner - Product Code, Community Code Aggregated

yrpc_ca <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrpc-ca?y=1962&r=usa&p=mex&o=17"
))

yrpc_ca
```

Columns definition:

* `community_code`: Unofficial Harvard CID community code (e.g. according to the \code{communities} table in the API, 17 stands for "Transportation")

### YRPC-GCA (Year - Reporter - Partner - Product Code, Group Code and Community Code Aggregated)

The applicable filters here are year, reporter, partner and community code. Here the community code is just an aggregation over both product code and group code.

```{r yrpc-gca, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Reporter - Partner - Product Code, Community Code Aggregated

yrpc_gca <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrpc-gca?y=1962&r=usa&p=mex&o=17"
))

yrpc_gca
```

### YRC (Year, Reporter and Product Code)

The only applicable filter is by year, reporter, product code and (optionally) product code length.

```{r yrc, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Reporter - Product Code

yrc <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrc?y=1962&r=chl"
))

yrc
```

Columns definition:

* `export_rca`:  Balassa Index or [Revealed Comparative Advantage](https://docs.tradestatistics.io/the-mathematics-of-economic-complexity.html#revealed-comparative-advantage-rca) of an exported product. 
* `import_rca`:  Balassa Index or [Revealed Comparative Advantage](https://docs.tradestatistics.io/the-mathematics-of-economic-complexity.html#revealed-comparative-advantage-rca) of and imported product. 

### YRP (Year, Reporter and Partner)

The only applicable filter is by year, reporter and partner.

```{r yrp, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Reporter - Partner
yrp <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrp?y=2018&r=chl&p=arg"
))

yrp
```

### YC (Year and Product Code)

The only applicable filter is by year, product and (optionally) product code length.

```{r yc, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Product Code
yc <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yc?y=2018&c=0101"
))

yc
```

Columns definition:

* `pci_fitness_method`: Product Complexity Index (PCI) computed by using the [Fitness Method](https://docs.tradestatistics.io/the-mathematics-of-economic-complexity.html#fitness-method).
* `pci_reflections_method`: Product Complexity Index (PCI) computed by using the [Reflections Method](https://docs.tradestatistics.io/the-mathematics-of-economic-complexity.html#reflections-method).
* `pci_eigenvalues_method`: Product Complexity Index (PCI) computed by using the [Eigenvalues Method](https://docs.tradestatistics.io/the-mathematics-of-economic-complexity.html#eigenvalues-method).
* `pci_rank_*_method`: The rank of a product given its PCI (e.g. the highest PCI obtains the #1)

### YR (Year and Reporter)

The only applicable filter is by year and reporter.

```{r yr, warning=FALSE, message=FALSE, cache = TRUE}
## Year - Reporter
yr <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yr?y=2018&r=chl"
))

yr
```

Columns definition:

* `cci_fitness_method`: Country Complexity Index (CCI) computed by using the [Fitness Method](https://docs.tradestatistics.io/the-mathematics-of-economic-complexity.html#fitness-method).
* `cci_reflections_method`: Country Complexity Index (CCI) computed by using the [Reflections Method](https://docs.tradestatistics.io/the-mathematics-of-economic-complexity.html#reflections-method).
* `cci_eigenvalues_method`: Country Complexity Index (CCI) computed by using the [Eigenvalues Method](https://docs.tradestatistics.io/the-mathematics-of-economic-complexity.html#eigenvalues-method).
* `cci_rank_*_method`: The rank of a product given its CCI (e.g. the highest CCI obtains the #1)

### Other group/community aggregated tables

As you might notice in [api.tradestatistics.io/tables](https://api.tradestatistics.io/tables), there are more tables:

* yrc-ga
* yrc-ca
* yrc-gca
* yr-short
* yr-ga
* yr-ca

These tables follow the same parameters as the examples above.

### Country rankings

The only applicable filter is by year.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Country rankings
country_rankings <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/country_rankings?y=2018"
))
```

### Product rankings

The only applicable filter is by year.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Product rankings
product_rankings <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/product_rankings?y=2018"
))
```

## R Package

To ease API using, we provide an [R Package](https://ropensci.github.io/tradestatistics/). This package is a part of [ROpenSci](https://ropensci.org/) and its documentation is available on a separate [pkgdown site](https://ropensci.github.io/tradestatistics/).

Here's what the package does:

```{r, echo = F, fig.cap="R package flow"}
knitr::include_graphics("fig/data-diagram.svg")
```

## Dashboard (beta)

To ease API using, we provide a [Shiny Dashboard](https://shiny.tradestatistics.io/) that is still under improvements.

## RDS datasets

Please check the [md5sums](https://docs.tradestatistics.io/direct-download/md5sums.txt) to verify data integrity after downloading.

```{r, echo = FALSE, message = F}
library(readr)
library(tidyr)

files <- list.files("docs/direct-download", pattern = "rds", full.names = TRUE)
urls <- str_replace(files, "docs/direct-download/", "https://docs.tradestatistics.io/direct-download/")

md5sums <- read_fwf("docs/direct-download/md5sums.txt", fwf_widths(c(32, 15), c("md5sum", "file"))) %>% mutate(file = str_trim(file))

tibble(
  `Year` = paste0("[", 1962:2018, "](", urls, ")"),
  `Last updated` = "2020-02-11",
  `File size (MB)` = round(file.size(files) / 1024^2, 2),
  `MD5 sum` = md5sums$md5sum
) %>%
  kable()
```
