# Using our datasets

## Compressed data

There is a special consideration you should have with our datasets, and is that you should always read the trade values as a numeric column and the commodity codes as a character column.

Different R packages, and statistical software in general, have includeded functions to autodetect column types. In our experience, that can read commodity codes as integers and that would ignore leading zeroes in commodity codes. The same applies to trade values that can be detected as integers after the program reads the first $n$ rows, and that would lead to read large values incorrectly due to integer class maximum value of 2,147,483,647.

As an example, let's read 1962 data:

```{r, warning=FALSE, message=FALSE}
# packages

library(data.table)
library(dplyr)
library(stringr)
library(janitor)

# custom functions

messageline <- function() {
  message(rep("-", 60))
}

fread2 <- function(file, select = NULL, character = NULL, numeric = NULL) {
  messageline()
  message("function fread2")
  message("file: ", file)
  
  if(str_sub(file, start = -2) == "gz") {
    d <- fread(
      cmd = paste("zcat", file),
      select = select,
      colClasses = list(
        character = character,
        numeric = numeric
      )
    ) %>%
      as_tibble() %>%
      clean_names()
  } else {
    d <- fread(
      input = file,
      select = select,
      colClasses = list(
        character = character,
        numeric = numeric
      )
    ) %>%
      as_tibble() %>%
      clean_names()
  }
  
  return(d)
}

# download data

url_1962 <- "https://tradestatistics.io/data/06-tables/hs-rev2007/1-yrpc/yrpc-1962.csv.gz"
gz_1962 <- "yrpc-2016.csv.gz"

if (!file.exists(gz_1962)) {
  try(download.file(url_1962, gz_1962))
}

# read data

data_1962 <- fread2(
      "yrpc-2016.csv.gz",
      character = "commodity_code",
      numeric = c(
        "export_value_usd",
        "import_value_usd",
        "export_value_usd_change_1_year",
        "export_value_usd_change_5_years",
        "export_value_usd_percentage_change_1_year",
        "export_value_usd_percentage_change_5_years",
        "import_value_usd_change_1_year",
        "import_value_usd_change_5_years",
        "import_value_usd_percentage_change_1_year",
        "import_value_usd_percentage_change_5_years"
      )
    )

data_1962
```

## API

By using the API you can obtain the same result as in section \@ref(compressed-data), but in a simpler way: 

```{r, warning = FALSE, message=FALSE, cache = TRUE}
library(jsonlite)

data_1962 <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrpc?y=1962&r=all&p=all&l=4"
))

data_1962
```

Because the API uses PostgreSQL queries that are converted to JSON, the columns are always read correctly with `jsonlite` R package.

The last API call makes sense with respect to `y`, `r` and `p`, but `l` is a requiered parameter that will be detailed in the forthcoming "Data" subsection.

### Metadata

The advantage of the API over https download is that you can filter what to obtain and also access some additional tables:

```{r, warning=FALSE, message=FALSE, cache = TRUE}
## Countries (no filter)
countries <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/countries"
))

## Products (no filter)
products <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/products"
))
```

```{r}
countries
```

```{r}
products
```

Please notice that these tables include some aliases. 

`countries` includes some meta-codes, `c-xx` where `xx` must the first two letters of a continent and `all`, this is:

```{r, echo = FALSE}
library(knitr)

countries %>% 
  filter(str_sub(country_iso, 1, 2) == "c-" | country_iso == "all") %>% 
  select(country_iso, country_name_english) %>% 
  rename(
    alias = country_iso,
    meaning = country_name_english
  ) %>% 
  kable()
```

`products` also includes some meta-codes, `xx` for the first two digits of a code and those digits are the product group and `all`, this is:

```{r, echo = FALSE}
products %>% 
  filter(str_length(commodity_code) %in% 2:3) %>% 
  select(commodity_code, product_fullname_english) %>% 
  rename(
    alias = commodity_code,
    meaning = product_fullname_english
  ) %>% 
  kable()
```

### Data

#### General parameters

The tables provided withing our API contain at least one of these fields:

* Year (`y`) 
* Reporter ISO (`r`)
* Partner ISO (`p`)
* Commodity code (`c`)

The most detailed table is `yrpc` that contains all bilateral flows at product level.

With respect to `y` you can pass any integer in $[1962,2016]$.

Both `r` and `p` accept any valid ISO code or alias contained in the [countries](https://api.tradestatistics.io/countries) table. For example, both `chl` (valid ISO code) and `c-am` (continent Americas, an alias) are valid API filtering parameters.

`c` takes any valid commodity code or alias from the [products](https://api.tradestatistics.io/products). For example, both `0101` (valid HS product code) and `01` (valid HS group code) are valid API filtering parameters.

In addition to `y, r, p, c` parameters, the length (`l`) parameter allows efficient queries provided our data contains both 4 and 6 digits long commodity codes. Because 4 digits code contain 6 digits codes, our approach is to allow the user to use `l=4`, `l=6` or `l=all` to provide just the requested data.

By default the API takes `c = "all"` and `l = 4` as defaults.

You can always skip `c` or `l`, but `y`, `r` and `p` are requiered to return data.

#### Available reporters

The only applicable filter is by year.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Available reporters (filter by year)
reporters <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/reporters?y=2015"
))
```

#### Country rankings

The only applicable filter is by year.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Country rankings (filter by year)
country_rankings <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/country_rankings?y=2015"
))
```

#### Product rankings

The only applicable filter is by year.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Product rankings (filter by year)
product_rankings <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/product_rankings?y=2015"
))
```

#### Year - Reporter - Partner - Commodity

You can filter by year, reporter and partner.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Reporter - Partner - Commodity (filter by year, reporter and partner)

## filter by commodity length (parameter `l`)
yrpc_1 <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrpc?y=2015&r=chl&p=arg&l=4"
))

yrpc_1

## filter by commodity group (parameter `c`)
yrpc_2 <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrpc?y=2015&r=chl&p=arg&c=01"
))

yrpc_2
```

Some columns requiere an explanation:

* `commodity_code`: HS07 product codes (e.g. according to the \code{products} table within this package, 0101 stands for "Horses, etc.")
* `commodity_code_length`: How many digits does `commodity_code` contain, this can be useful to filter by depth when using HS codes (HS 6 digits is a more detailed version of HS 4 digits, and therefore you don't have to sum both or you'll be counting exports/imports twice)
* `group_code`: International categorization of group products defined after product ID
* `group_name`: English name corresponding to `group_id`
* `export_value_usd`: Exports measured in nominal United States Dollars (USD)
* `import_value_usd`: Imports measured in nominal United States Dollars (USD)
* `export_value_usd_percentage_change_1_year`: Nominal increase/decrease in exports measured as percentage with respect to last year
* `export_value_usd_percentage_change_5_years`: Nominal increase/decrease in exports measured as percentage with respect to five years ago
* `export_value_usd_change_1_year`: Nominal increase/decrease in exports measured in USD with respect to last year
* `export_value_usd_change_5_years`: Nominal increase/decrease in exports measured in USD with respect to five years ago

#### Year - Reporter - Commodity

The only applicable filter is by year and reporter.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Reporter - Commodity (filter by year and reporter)

## filter by reporter ISO (parameter `r`)
yrc_1 <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrc?y=2015&r=chl&l=4"
))

yrc_1

## filter by reporter alias (also parameter `r`)
yrc_2 <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrc?y=2015&r=c-am&l=4"
))

yrc_2
```

Here the `export_rca*` and `import_rca*` fields contain the Revealed Comparative Advantage (RCA) of an exported product with respect to all the products with the same number of digits. The definition of RCA is detailed on [Open Trade Statistics Documentation](https://tradestatistics.github.io/documentation/).

#### Year - Reporter - Partner

The only applicable filter is by year, reporter and partner.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Reporter - Partner (filter by year, reporter and partner)
yrp <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yrp?y=2015&r=chl&p=arg"
))
```

#### Year - Commodity

The only applicable filter is by year and commodity.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
# Year - Commodity (filter by year)
yc <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yc?y=2015&c=0101"
))
```

Let's explore the first rows of `yr`:

```{r}
yc
```

Here some fields deserve an explanation:

* `pci_4_digits_commodity_code`: Product Complexity Index (PCI) which is detailed on [Open Trade Statistics Documentation](https://tradestatistics.github.io/documentation/). This index is built by using just four digits commodity codes.
* `pci_6_digits_commodity_code`: Similar to the previous field but built by using just six digits commodity codes.
* `pci_rank_4_digits_commodity_code`: The rank of a product given its PCI (e.g. the highest PCI obtains the #1)
* `pci_rank_4_digits_commodity_code_delta_1_year`: How many places a country increased or decreased with respect to last year

#### Year - Reporter

The only applicable filter is by year and reporter.

```{r, warning=FALSE, message=FALSE, cache = TRUE}
## Year - Reporter (filter by year and reporter)
yr <- as_tibble(fromJSON(
  "https://api.tradestatistics.io/yr?y=2015&r=chl"
))
```

Let's explore the first rows of `yr`:

```{r}
yr
```

Some fields here require more detail:

* `eci_4_digits_commodity_code`: Economic Complexity Index (ECI) which is detailed on [Open Trade Statistics Documentation](https://tradestatistics.github.io/documentation/). This index is built by using just four digits commodity codes.
* `eci_rank_4_digits_commodity_code`: The rank of a country given its ECI (e.g. the highest ECI obtains the #1)
* `eci_rank_4_digits_commodity_code_delta_1_year`: How many places a country increased or decreased with respect to last year
